---
title: "Probability and diagnostic testing"
format: 
  html:
    code-fold: false
    code-tools: true
editor: source
---

In this page you will find some examples from the lecture. You are not required to do your exercises with R, however R can be used as a calculator. 

* Sensitivity = TP / P = TP / (TP+FN)
* Specificity = TN / N = TN / (TN+FP)
* Positive predictive value PPV = TP / test positive = TP / (TP + FP)

In addition,

* false positive rate (FPR) is 1 - specificity
* false negative rate (FNR) is 1 - sensitivity

When you have prevalence of the disease, the positive and negative predictive value PPV, NPV are

$$PPV = \frac{sens \times prev}{sens \times prev + (1-spec) \times (1 - prev)}$$

$$NPV = \frac{spec \times prev}{(1-sens) \ times prev + spec \times (1-prev)}$$

## Mammography example

In the mammography dataset, we are given the counts of patients that fall into one of the four categories: 

* 22 tested cancer really have cancer
* 331 tested no cancer really do not have cancer
* 16 tested cancer, but do not have cancer
* 3 tested no cancer, but really have cancer

According to the terminology introduced in class, we can create four variables to store the data.

```{r}
#| label: mamm1
#| warning: false
#| echo: true

tp <- 22
tn <- 331
fp <- 16
fn <- 3
```

We can find how many patients really have cancer, or not have cancer. Here **positive** and **negative** refer to the actual disease status, not thest test result.

```{r}
#| label: mamm2
#| warning: false
#| echo: true

# positive means positive condition - real disease status
positives <- tp + fn
positives
negatives <- tn + fp
negatives
```

Now find the **sensitivity, specificity, positive predictive value**. 

```{r}
#| label: mamm3
#| warning: false
#| echo: true

# sensitivity: tp / positives
tp/positives

# specificity: tn / negatives
tn/negatives

# positive predictive value ppv
# tp / positive test
tp / (tp+fp)
```


## HIV example

The HIV example from the class demonstrates the how prevalence affects metrics of diagnostic tests. 

To start, we set prevalence to 0.001.


```{r}
#| label: hiv1
#| warning: false
#| echo: true

# prevalence 0.1%
# other values
# prevalence <- 0.01
# prevalence <- 0.1
# prevalence <- 0.0001
prevalence <- 0.001
```

We were given some terms that are not sensitivity and specifity, but it is straightfoward to translate into terms we know.

```{r}
#| label: hiv2
#| warning: false
#| echo: true

# false positive rate 0.2%
# fpr is 1-specificity
specificity <- 1-0.002
specificity
# false negative rate 2%
# fnr is 1-sensitivity
sensitivity <- 1-0.02
sensitivity
```

Find positive predictive value from the formula.

```{r}
#| label: hiv3
#| warning: false
#| echo: true

a <- sensitivity * prevalence
b <- sensitivity * prevalence + (1-specificity) * (1-prevalence)
a/b

# you can skip the step where you name a and b
(sensitivity * prevalence) / (sensitivity * prevalence + (1-specificity) * (1-prevalence))
```