---
title: "Survival analysis"
description: "Kaplan-Meier curve, log rank tests"
format: 
  html:
    code-fold: false
    code-tools: true
editor: source
---

Datasets

* Exercise 1: `melanoma`([rda link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/melanoma.rda), [csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/melanoma.csv))
* Exercise 2: `liggetid` ([csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/melanoma.csv))

[R script](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/code/4_ttest.R)

------------------




# Exercise 1: Melanoma 

The data concern 205 patients with malignant melanoma who were operated at the university hospital in Odense, Denmark, in the period 1962-77. The patients were followed up to death or censored at the end of the study. We shall study the effect on survival of the patient’s gender and tumor thickness.

The variables included in the dataset are:

* status (1=death from disease, 2=censored, 4=death from other causes) 
* lifetime (years) from operation
* ulceration of tumor (1=yes, 2=no)
* tumor thickness in 1/100 mm
* gender (f=1, m=2)
* age at operation in years
* grouped tumor thickness (1: 0-1 mm, 2: 2-5 mm, 3: 5+ mm) 
* logarithm of tumor thickness


First we load the dataset, and check what columns are there.

```{r}
#| label: melanoma-1
#| warning: false
#| echo: true
library(survival)

melanoma <- haven::read_dta('data/melanoma.dta')

head(melanoma)
colnames(melanoma)
```

### 1a) 

The “event of interest” is defined as “death from disease”. Pre-process the variables that R will use for basic survival analysis as shown in class.

We only need "death from melanoma" or else, hence we need to recode the `status` variable which originally had three categories.

```{r}
#| label: melanoma-a1
#| warning: false
#| echo: true

# if status == 1, code 1; otherwise, code 0
# this indicates whether deaths from melanoma
death <- ifelse(melanoma$status == 1, 1, 0)

# check if the labels are correct
table(death)
table(melanoma$status)

```

You can see that the category `melanoma$status == 1` should match `death == 1`, and `melanoma$status` of 2 and 4 are merged together and coded as `death == 0`. 

### 1b) 

Make a Kaplan-Meier plot of the survival curve. What are the estimated probabilities of surviving 1 year, 2 years, 5 years?


```{r}
#| label: melanoma-b1
#| warning: false
#| echo: true

# take lifetime variable
lifetime <- melanoma$lifetime


km_fit <- survfit(Surv(lifetime, death) ~ 1)
plot(km_fit)

# add title and text
title(main = 'Kaplan-Meier survival estimate', 
      xlab = 'Time', 
      ylab = 'Survival probability')

```


Now we check only time 1, 2, 5. 

```{r}
#| label: melanoma-b2
#| warning: false
#| echo: true

# time 1, 2, 5
tme <- c(1, 2, 5)
summary(km_fit, times = tme)
```


### 1c)

Compare survival according to gender. Test the difference with a logrank test.


```{r}
#| label: melanoma-c1
#| warning: false
#| echo: true

# gender == 1: female; gender == 2: male
gender <- melanoma$gender


km_fit_gender <- survfit(Surv(lifetime, death) ~ gender)
plot(km_fit_gender, col = c('blue', 'red'))
title(main = 'Kaplan-Meier survival estimate: stratify by gender', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'bottomleft', 
       legend = c('Female', 'Male'), 
       lty = c('solid', 'solid'), 
       col = c('blue','red'))

```

Now we test the difference with log-rank test

```{r}
#| label: melanoma-c2
#| warning: false
#| echo: true

# test difference with logrank test 
survdiff(Surv(lifetime, death) ~ gender)
```


### 1d) 

Study survival according to grouped tumor thickness by comparing survival curves in each group and test with a logrank test.

First we do some data processing

```{r}
#| label: melanoma-d1
#| warning: false
#| echo: true

# check the original data type 
melanoma$grouped_tumor_thickness

# you can code them so that the data is more readable
grouped_tumor_thickness <- factor(melanoma$grouped_tumor_thickness, 
                           levels = c('1','2', '3'),
                           labels = c('0-2 mm', '2-5 mm', '5+ mm'))

# check if it is coded correctly
table(grouped_tumor_thickness)
table(melanoma$grouped_tumor_thickness)

```


Now we plot the KM plot, and carry out a log-rank test
```{r}
#| label: melanoma-d2
#| warning: false
#| echo: true

km_fit_tumor <- survfit(Surv(lifetime, death) ~ grouped_tumor_thickness)
plot(km_fit_tumor, col = c('blue', 'red', 'forestgreen'))

title(main = 'Kaplan-Meier survival estimate: stratify by tumor thickness', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'bottomleft', 
        legend = c('0-2 mm', '2-5 mm', '5+ mm'), 
        lty = c('solid', 'solid', 'solid'), 
        col = c('blue','red', 'forestgreen'))


# test difference with logrank test 
survdiff(Surv(lifetime, death) ~ grouped_tumor_thickness)

```




# Exercise 2: Length of hospital stay 

The data was collected at the Geriatric Department at Ullevål Sykehus. Below is a description of the data set `liggetid`. The file includes the following variables:

- Year of birth (faar)
- Month of birth (fmaan)
- Day of birth (fdag)
- Year of hospital admission (innaar)
- Month of admission (innmaan)
- Day of admission (inndag)
- Year of discharge from hospital (utaar)
- Month of discharge (utmaan)
- Day of discharge (utdag)
- Gender, where 1 = male and 0 = female (kjoenn)
- Admission from, where 1 = home, 2 = Div. of Medicine, 3 =
Div. of Surgery, 4 = Other division, 5 = Other hospital, 6 =
Nursing home (kom_fra)
- Stroke, where 1 = yes, 0 = no (slag)
- Age (alder)
- Hospital stay, in days (liggetid)
- Logarithm of hospital stay (lnliggti)
- Comes from Div. of Medicine (kom_fra2)
- Comes from Div. of Surgery (kom_fra3)
- Comes from Other division (kom_fra4)
- Comes from Other hospital (kom_fra5)
- Comes from Nursing home (kom_fra6)
- Censoring variable (censor)

The variable `liggetid` time is calculated from the `innaar`, `innmaan`, `inndag`, `utaar`, `utmaan` and `utdag` variables. 


```{r}
#| label: liggetid
#| warning: false
#| echo: true

liggetid <- haven::read_dta('data/liggetid.dta')
head(liggetid)
colnames(liggetid)

# all status are 1 
status <- liggetid$status
```


### 2a) 

Analyze the relationship between the variables `liggetid` and `kjoenn` with a Kaplan-Meier plot. Test the difference with a log-rank test.


```{r}
#| label: liggetid-a1
#| warning: false
#| echo: true

# length of stay (los) vs gender
los <- liggetid$liggetid
head(los)

# can choose to recode; or leave as it is;
# gender == 0: female; gender == 1: male
gender <- liggetid$kjoenn
head(gender)

# fit km 
km_liggetid_gender <- survfit(Surv(los, status) ~ gender)

plot(km_liggetid_gender, col = c('blue', 'red'))

title(main = 'Kaplan-Meier survival estimate: stratify by gender', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'topright', 
        legend = c('Female', 'Male'), 
        lty = c('solid', 'solid'), 
        col = c('blue','red'))

```

Now we do log-rank test

```{r}
#| label: liggetid-a2
#| warning: false
#| echo: true

# log rank test 
survdiff(Surv(los, status) ~ gender)
```


### 2b) 

Analyze the relationship between the variables `liggetid` and `slag` with a Kaplan-Meier plot. Test the difference with a log-rank test.


```{r}
#| label: liggetid-b1
#| warning: false
#| echo: true


# length of stay vs stroke
# slag == 1: yes, slag == 2: no
stroke <- liggetid$slag

# fit km 
km_liggetid_stroke <- survfit(Surv(los, status) ~ stroke)

plot(km_liggetid_stroke, col = c('blue', 'red'))

title(main = 'Kaplan-Meier survival estimate: stratify by stroke', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'topright', 
        legend = c('Stroke: yes', 'Stroke: no'), 
        lty = c('solid', 'solid'), 
        col = c('blue','red'))

```


```{r}
#| label: liggetid-b2
#| warning: false
#| echo: true

# log rank test
survdiff(Surv(los, status) ~ stroke)
```