---
title: "Non-parametric tests"
description: "Wilcoxon signed rank, rank sum test"
format: 
  html:
    code-fold: false
    code-tools: true
editor: source
---

# Short summary


# Examples (exercises with solution)


## Exercise 1 (lung function)

Lung function has been measured on 106 medical students. Peak expiratory flow rate (PEF, measured in liters per minute) was measured three times in a sittinng position, and three times in a standing position.

The variables are 

* Age (years)
* Gender (female, 2 male)
* Height (cm)
* Weight (kg)
* PEF measured three times in a sitting position (pefsit1, pefsit2, pefsit3)
* PEF measured three times in a standing position (pefsta1, pefsta2, pefsta3)
* Mean of the three measurements made in a sitting position (pefsitm)
* Mean of the three measurements made in a standing position (pefstam)
* Mean of all six PEF values (pefmean)


#### 1a) 

We shall consider the difference betweenn PEF measured in a sitting position and in a standing position, i.e. we look at `diff_pef = pefstam - pefsitm`.  First compute this variable. 


```{r}
#| label: nonpara-1a-1
#| echo: true
#| eval: true

lung_data <- read.csv('data/PEFH98-english.csv')
head(lung_data, 3)

# compute diff_pef
diff_pef <- lung_data$pefstam - lung_data$pefsitm
```


#### 1b) 

Make a histogram and Q-Q plot for the differences. 

```{r}
#| label: nonpara-1b-1
#| echo: true
#| eval: true

par(mfrow = c(1, 2))
hist(diff_pef)

# qqplot 
qqnorm(diff_pef)
qqline(diff_pef)
```


#### 1c) 

Do a non-parametric test to decide whether `pefsitm` and `pefstam` are significantly different.

```{r}
#| label: nonpara-1c-1
#| echo: true
#| eval: true

# one sample (signed rank test)
wilcox.test(diff_pef)
```



```{r}
#| label: nonpara-1c-2
#| echo: true
#| eval: true

# alternatively
pefstam <- lung_data$pefstam
pefsitm <- lung_data$pefsitm

wilcox.test(pefstam, pefsitm, paired = T) # paired 
```


```{r}
#| label: nonpara-1c-3
#| echo: true
#| eval: true

# the data is not severly skewed, can also use t-test 
t.test(pefstam, pefsitm, paired = T)
```




## Exercise 2 (length of hospital stay)

The data was collected at the Geriatric Department at Ullevål Sykehus. Below is a description of the data set `liggetid`. The file includes the following variables:

- Year of birth (faar)
- Month of birth (fmaan)
- Day of birth (fdag)
- Year of hospital admission (innaar)
- Month of admission (innmaan)
- Day of admission (inndag)
- Year of discharge from hospital (utaar)
- Month of discharge (utmaan)
- Day of discharge (utdag)
- Gender, where 1 = male and 0 = female (kjoenn)
- Admission from, where 1 = home, 2 = Div. of Medicine, 3 =
Div. of Surgery, 4 = Other division, 5 = Other hospital, 6 =
Nursing home (kom_fra)
- Stroke, where 1 = yes, 0 = no (slag)
- Age (alder)
- Hospital stay, in days (liggetid)
- Logarithm of hospital stay (lnliggti)
- Comes from Div. of Medicine (kom_fra2)
- Comes from Div. of Surgery (kom_fra3)
- Comes from Other division (kom_fra4)
- Comes from Other hospital (kom_fra5)
- Comes from Nursing home (kom_fra6)
- Censoring variable (censor)

The variable `liggetid` time is calculated from the `innaar`, `innmaan`, `inndag`, `utaar`, `utmaan` and `utdag` variables. 



#### 2a)

We want to compare the **length of hospital stay** for patients with and without **stroke**. First get a better understanding of the data. Start with computing some summary (descriptive) statistics for these two variables: 

* **mean, median, min, max, 75-th percentile** for length of hospital stay (`liggetid`);
* **count** of how many patient have stroke, and do not have stroke (`slag == 1` or `0`)

Visualize the variable `liggetid` using histogram and Q-Q plot, is the requirement for t-test met?


```{r}
#| label: nonpara-2a-1
#| echo: true
#| eval: true

liggetid <- read.csv('data/liggetid.csv')
head(liggetid, 3)
```



```{r}
#| label: nonpara-2a-2
#| echo: true
#| eval: true

# liggetid (length of stay: los)
los <- liggetid$liggetid
stroke <- liggetid$slag

summary(los)

table(stroke)
summary(stroke) # this shows 85 missing
sum(is.na(stroke)) # 85
```



```{r}
#| label: nonpara-2a-3
#| echo: true
#| eval: true

par(mfrow = c(1, 2))
# visualize length of hospital stay (los)
hist(los, breaks = 25)
# check normality
qqnorm(los, pch = 20, main = 'Q-Q plot for length of stay')
qqline(los, col = 'red', lwd = 2)
```




#### 2b) 

Since we want to compare the length of hospital stay for two groups, we need to separate the length of stay data into two groups, based on whether patients have stroke or not. Create two variables, `ligt_s1` and `ligt_s0`. 

Do an appropriate non-parametric test. What is the difference between this test and the one you did in Exercise 1?


::: {.callout-tip collapse="true"}
## Separate one variable via index and `which()`

Previously, we showed how to separate one variable based on another,

* Exercise 2c) in [EDA I](https://ocbe-uio.github.io/teaching_mf9130e/lab/lab_eda_part1.html)
* Exercise 2a) in [Categorical data analysis](https://ocbe-uio.github.io/teaching_mf9130e/lab/lab_categorical.html)

You can try to use this method here as well; but you might notice that there are some NA in the results. This is because when you do `stroke == 0` on NA, instead of giving you `T` or `F`, it gives `NA`; so the result of filtering `liggetid` keeps both `T` and `NA`. 

These NAs can be removed post-hoc; but a clearer solution is to set 2 conditions inside the square bracket, and combine with `&`. 

* `!is.na(stroke)` means stroke variable not NA
* `stroke == 1` means stroke variable equals to 1
* `which()` identifies the indices for stroke where the conditions were met, e.g. the 4th, 22th elements.

:::


```{r}
#| label: nonpara-2b-1
#| echo: true
#| eval: true


# separate los 
ligt_s1 <- los[which(!is.na(stroke) & stroke == 1)]
ligt_s0 <- los[which(!is.na(stroke) & stroke == 0)]

# mean, median 
c(mean(ligt_s1), median(ligt_s1))
c(mean(ligt_s0), median(ligt_s0))
```



```{r}
#| label: nonpara-2b-2
#| echo: true
#| eval: true

# visualize
par(mfrow = c(2,2))

# s1
hist(ligt_s1, main = 'Length of hospital stay, stroke yes', 
     xlab = 'mean: 145; median: 58', 
     xlim = c(0, 1300))
abline(v = mean(ligt_s1), col = 'blue', lwd = 2)
abline(v = median(ligt_s1), col = 'blue', lwd = 2, lty = 'dashed')
# qqplot
qqnorm(ligt_s1, pch = 20, main = 'Q-Q plot for length of stay, stroke yes')
qqline(ligt_s1, col = 'red', lwd = 2)

# s0
hist(ligt_s0, main = 'Length of hospital stay, stroke no', 
     xlab = 'mean: 116; median: 40', 
     xlim = c(0, 1300))
abline(v = mean(ligt_s0), col = 'blue', lwd = 2)
abline(v = median(ligt_s0), col = 'blue', lwd = 2, lty = 'dashed')
# qqplot
qqnorm(ligt_s0, pch = 20, main = 'Q-Q plot for length of stay, stroke no')
qqline(ligt_s0, col = 'red', lwd = 2)
```


```{r}
#| label: nonpara-2b-3
#| echo: true
#| eval: true

# non parametric test
# not matched; independent samples
# wilcoxon rank sum 
wilcox.test(ligt_s1, ligt_s0)

```

::: {.callout-caution}
## Compare with t-test

If you do a t-test (inappropriately) instead of Mann-Whitney U test, you'll have different results!

:::

```{r}
#| label: nonpara-2b-4
#| echo: true
#| eval: true

# t-test on skewed data
t.test(ligt_s1, ligt_s0)
```


#### 2c)


Do the same analysis with a t-test on the logarithm-transformed data. The transformed variable already exists in the data set as `lnliggti`.


```{r}
#| label: nonpara-2c-1
#| echo: true
#| eval: true


# use log transformed 
loglos <- liggetid$lnliggti
hist(loglos, main = 'Log transformed length of stay')
qqnorm(loglos, pch = 20, main = 'Q-Q plot for log(length of stay)')
qqline(loglos, col = 'red', lwd = 2)

# separate data
logligt_s1 <- loglos[which(!is.na(stroke) & stroke == 1)]
logligt_s0 <- loglos[which(!is.na(stroke) & stroke == 0)]

t.test(logligt_s1, logligt_s0, paired = F)
```



## Exercise 3 (antibody)

(Baker et al., 1980) We have data on concentration of antibody of Type III Group B Streptococcus (GBS) in 20 volunteers before and after immunization. The dataset is `antibody`. 

#### 3a) 

The comparison of the antibody levels was summarized in the report of this study as '$t = 1.8; P > 0.05$'. Comment on the result. 


```{r}
#| label: nonpara-3a-1
#| echo: true
#| eval: true

antibody <- read.csv('data/antibody.csv')
head(antibody)

before <- antibody$before
after <- antibody$after
```

Compute some summary statistics for the two variables, and then reproduce the result by the t-test.

```{r}
#| label: nonpara-3a-2
#| echo: true
#| eval: true

# some summary stat of the data
summary(before)
summary(after)

# reproduce the result from t-test
t.test(antibody$before, antibody$after, paired = T)
```

#### (Optional)

We can visually get an understanding of whether this result makes sense.

```{r}
#| label: nonpara-3a-3
#| echo: true
#| eval: true

# long format data (requires tidyr package)
antibody_longdata <- tidyr::pivot_longer(antibody, 
                                         cols = tidyr::everything(), 
                                         names_to = 'group', 
                                         values_to = 'measurements')
head(antibody_longdata)
```



```{r}
#| label: nonpara-3a-4
#| echo: true
#| eval: true

par(mfrow = c(1, 2))
boxplot(measurements ~ group, 
        data = antibody_longdata, 
        horizontal = T, 
        main = 'Before and after measurements')

# limit the y axis (horizontal x axis)
boxplot(measurements ~ group, 
        data = antibody_longdata, 
        horizontal = T, 
        ylim = c(0, 4),
        main = 'Limit the measurement range')

```


```{r}
#| label: nonpara-3a-5
#| echo: true
#| eval: true

par(mfrow = c(2,2))
# histogram 
hist(before, main = 'Measurements: before')
qqnorm(before, pch = 20, main = 'Q-Q plot for measurements: before')
qqline(before, col = 'red', lwd = 2)

hist(after, main = 'Measurements: after')
qqnorm(after, pch = 20, main = 'Q-Q plot for measurements: after')
qqline(after, col = 'red', lwd = 2)
```


```{r}
#| label: nonpara-3a-6
#| echo: true
#| eval: true

# histogram for the difference s
hist(after - before, main = 'Difference (after - before)')
```

#### 3b)

Analyse the data with an appropriate test. Justify your choice. 

```{r}
#| label: nonpara-3b-1
#| echo: true
#| eval: true

# matched data (paired)
# one-sample test (signed rank)

wilcox.test(before, after, paired = T)
# the error means there are ties (no change)
```

::: {.callout-caution}
## Paired vs independent data

You can try to do a test treating the data as independent. The conclusion could be different! Be careful with what you are testing.

:::

```{r}
#| label: nonpara-3b-2
#| echo: true
#| eval: true

# what happens if you use independent two-samples test?
wilcox.test(before, after, paired = F) # p = 0.14
```

## Exercise 4 (nausea) 

(Williams et al., 1989) Patients receiving chemotherapy as outpatients were randomized to receive either an active antiemetic treatment or placebo. The following table shows measurements (in mm) on a 100mm linear analogue self-assessment scale for nausea. The dataset is `nausea`. 

* group: `1` is active (treatment), and `2` is placebo.

#### 4a)

Choose an appropriate test for comparing the treatment and placebo groups. Justify your choice. 

```{r}
#| label: nonpara-4a-1
#| echo: true
#| eval: true

nausea <- read.csv('data/nausea.csv')
head(nausea)

treatment <- nausea$value[nausea$group == 1]
placebo <- nausea$value[nausea$group == 2]

summary(treatment)
summary(placebo)
```


```{r}
#| label: nonpara-4a-2
#| echo: true
#| eval: true

# these samples are not matched: two sample independent test

# boxplot
boxplot(value ~ group, data = nausea, 
        main = 'Nausea: treatment vs placebo')

```


```{r}
#| label: nonpara-4a-3
#| echo: true
#| eval: true

wilcox.test(treatment, placebo, paired = F)
```