---
title: "Descriptive statistics"
description: "R lab 2"
format: 
  html:
    code-fold: false
    code-tools: true
editor: visual
---

## Association modeling (placeholder for the R labs)

In this part of the exercise, we model (on the log-scale) the association of miRNA espression on protein expression adjusting for the corresponding mRNA.

Investigate **miR-107** and **B-RAF** (Aure et al, 2015, Figure 2H)

```{r}
#| label: load-data-again
#| warning: false
#| echo: false

# need to load the data in this page again. DO NOT SHOW
mir <- read.table("data/miRNA-421x282.txt", header=T, sep="\t", dec=".")
rna <- read.table("data/mRNA-100x282.txt", header=T, sep="\t", dec=".")
prt <- read.table("data/prot-100x282.txt", header=T, sep="\t", dec=".")

mir <- as.matrix(mir)
rna <- as.matrix(rna)
prt <- as.matrix(prt)
```

```{r}
#| label: select-braf
#| warning: false
#| echo: true
 
prt.BRAF = prt[12,]
rna.BRAF = rna[12,]
mir.107 = mir[16,] 
```

### (a) Linear regression model

on the log-scale, Aure et al. 2015, equation (3)

```{r}
#| echo: true
#| eval: true

fitA <- lm(prt.BRAF ~ mir.107 + rna.BRAF)
summary(fitA)
```

::: callout-note
## Task 3

Repeat the lasso analysis, but this time do not penalise the `rna.BRAF` variable together with the mir variables.

Check out the information on the `penalty.factor` option in `?glmnet` to understand how.
:::
