---
title: "Introduction to R"
description: "Get started in Rstudio, navigation, data types, data import ant export"
format: 
  html:
    code-fold: false
    code-tools: false
editor: source
---

**Material on this page is under development. There could be errors. **

## R in a nutshell 

R is a language and environment for statistical computing, data analysis, visualisation and graphics and many more. It is a free and open source software, under the terms of GNU General Public License. 

R runs on a wide variety of platforms, including Windows, Linux and MacOS. 

The latest R version is 4.2.3 (2023-03-15), "Shortstop Beagle".

### RStudio and Posit

RStudio is a free and open-source IDE (integrated development environment). RStudio IDE is developed by Posit, previously RStudio PBC. 

It is possible to download RStudio to your laptop or desktop; or use RStudio Cloud (need internet connection).

**4 Panels, use screenshot**



### Getting help 

* Use Google well! There are a lot of discussions in forums such as StackOverflow
* Cheatsheets made by Posit and community
* Type `?functionName` in the console 
* Package vignettes (longer format of documentation)



# Basic commands 

### Assignment of object

r as calculator (**use console first; use script AFTER project and workspace mgmt!**)

```{r}
#| label: assign value
#| warning: false
#| echo: true

# assign values to variables a, b, c
# (comments start with #)
a <- 3
b <- 4
c <- 7

# calculate the mean
# output directly
(a+b+c)/3

# or, save into a new variable d
d <- (a+b+c)/3
d

```


# Functions and packages


**just at abstract level for now (in the lecture), students can learn after seeing many examples**

**in this document, would still be useful to include links so they can read on their own**



# Data import and export 

project management 

set work directory


import data of different formats (txt, csv, xlsx, rda...)

```{r}
#| label: load-data-penguins-mini
#| warning: false
#| echo: true

# need to load the data in this page again. DO NOT SHOW
penguins_mini <- read.csv('data/penguins_mini.csv', sep = ',')

penguins_mini
```




# Data structure (use `penguins_mini` data as example)

**Bec student background, it is not necessary to go into too much details**

**MUST: concept of scalar, vector, matrix and data frame** (dataset exploration put to descriptive and eda)

**MUST: numeric, factor, character, logical**







# Visualisation (only intro, details later)

### (base R)


### ggplot

Just some basics, since more specific types of viz will be shown in each section

::: callout-note
## Task 3

Try explore the penguins data
:::
