---
title: "Regression analysis"
description: "Linear regression, correlation"
format: 
  html:
    code-fold: false
    code-tools: true
editor: source
---


Datasets

* Exercise 1: `bp` ([rda link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/bp.rda), [csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/bp.csv))
* Exercise 2: `PEFH98-english` ([rda link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/PEFH98-english.rda), [csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/PEFH98-english.csv))
* Exercise 3: `liggetid` ([rda link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/liggetid.rda), [csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/liggetid.csv))


[R Script](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/code/8_linearreg.R)


------------------

# Examples (exercises with solution)


## Exercise 1 (blood pressure)

The dataset `bp` contains data on 20 healthy adults on two variables, `Age` and `Blood pressure`. We will explore the relationship between these two variables.


#### 1a)

Load the dataset. Find the correlation between age and blood pressure, and test if it is significant. Compute a 95% confidence interval for the regresion parameter.  

Also find the squared correlation coefficient between age and blood pressure. What does it mean?


```{r}
#| label: linear-1a-1
#| warning: false
#| echo: true

# load data
bp <- read.csv('data/bp.csv')
head(bp)
```


```{r}
#| label: linear-1a-2
#| warning: false
#| echo: true

# correlation age vs bp
cor(bp$Age, bp$bloodpressure)

# or,
cor(bp)

# 95% CI, p-value
cor.test(bp$Age, bp$bloodpressure) 
```



#### 1b)

What is the blood pressure for a person at age 40? For a person at age 75? Comment.

```{r}
#| label: linear-1b-1
#| warning: false
#| echo: true

# fit a linear regression model
model_age_bp <- lm(bloodpressure ~ Age, data = bp)
summary(model_age_bp)

# to predict (insert x), you need to put data in a data frame
predict(model_age_bp, 
        newdata = data.frame(Age = c(40, 75)), 
        interval = 'prediction')
```

Can also visualize the relationship.

```{r}
#| label: linear-1b-2
#| warning: false
#| echo: true

plot(x = bp$Age, y = bp$bloodpressure, 
     main = 'Age versus Blood Pressure', 
     xlab = 'Age', ylab = 'Blood pressure', 
     pch = 20, 
     xlim = c(15, 80))
# add the regression line on top
abline(lm(bloodpressure ~ Age, data = bp), 
       col = 'blue', lwd = 3)
```



## Exercise 2 (lung function)

Lung function has been measured on 106 medical students. Peak expiratory flow rate (PEF, measured in liters per minute) was measured three times in a sittinng position, and three times in a standing position.

The variables are 

* Age (years)
* Gender (female, male)
* Height (cm)
* Weight (kg)
* PEF measured three times in a sitting position (pefsit1, pefsit2, pefsit3)
* PEF measured three times in a standing position (pefsta1, pefsta2, pefsta3)
* Mean of the three measurements made in a sitting position (pefsitm)
* Mean of the three measurements made in a standing position (pefstam)
* Mean of all six PEF values (pefmean)


#### 2a) 

Make a scatter plot of `pefsit2` versus `pefsit1`; and make a separate scatter plot of `pefsit1` versus `weight`. Insert a regression line on top of the scatterplots. 


```{r}
#| label: linear-2a-1
#| warning: false
#| echo: true

lung_data <- read.csv('data/PEFH98-english.csv')
# head(lung_data)
# assign variables (not strictly necessary)
pefsit1 <- lung_data$pefsit1
pefsit2 <- lung_data$pefsit2
weight <- lung_data$weight

par(mfrow = c(1, 2)) # make plots in 1 row 2 col

# scatter plot: pefsit2 vs pefsit1
plot(x = pefsit1, y = pefsit2, 
     main = 'PEF sit1 vs PEF sit2', 
     xlab = 'pefsit1', ylab = 'pefsit2', 
     pch = 20)
abline(lm(pefsit2 ~ pefsit1, data = lung_data), 
       col = 'blue', lwd = 3)

# scatter plot: pefsit1 vs weight
plot(x = weight, y = pefsit1, 
     main = 'Weight vs pefsit1', 
     xlab = 'Weight', ylab = 'pefsit1', 
     pch = 20)
abline(lm(pefsit1 ~ weight, data = lung_data), 
       col = 'blue', lwd = 3)

```

#### 2b) 

Compute the correlation between `pefsit1` and `pefsit2`; and between `pefsit1` and `weight`. 

Why is the correlation between the first pair closer to 1 than the second pair?

(You can get the pair-wise correlation between many other pairs of variables using `cor(your_data)`). 

```{r}
#| label: linear-2b-1
#| warning: false
#| echo: true

cor(pefsit2, pefsit1) # need to remove NA here

which(is.na(pefsit1)) # no missing
which(is.na(pefsit2)) # 66th missing

# option 1: cor() removes NA for you 
# specify use complete observations
cor(pefsit2, pefsit1, use = 'complete.obs')

 
# option 2: you process (remove) the row of missing 
# from both variables (remove element 66)
pefsit2_narm <- pefsit2[!is.na(pefsit2)]
pefsit1_narm <- pefsit1[!is.na(pefsit2)]

# use pefsit2_narm instead of pefsit2 to compute cor
# should be the same
cor(pefsit2_narm, pefsit1_narm)
```

For `pefsit1` and `weight` it is straightforward, as there is not missing data. 

```{r}
#| label: linear-2b-2
#| warning: false
#| echo: true

# pefsit1, weight
cor(pefsit1, weight)
```

To compute the correlation between multiple pairs, you need to select a few variables first.

```{r}
#| label: linear-2b-3
#| warning: false
#| echo: true

# pairwise for multiple pairs 
# use age, height, weight, pefsit1, pefsit2, pefsit3, pefmean
# select a smaller dataset 
lungdata2 <- lung_data[, c('age', 'height', 'weight', 'pefsit1', 
                           'pefsit2', 'pefsit3', 'pefmean')]

head(lungdata2, 3)

# produce correlation matrix for all the variables here
# round(1.2345, digits = 2) gives 1.23
round(cor(lungdata2, use = 'complete.obs'), digits = 2)
```


#### 2c) 

Carry out two regression analysis: 

* `pefsit2` as dependent variable, `pefsit1` as independent variable;
* `pefsit1` as dependent variable, `weight` as independent variable.

Interpret the results in relation to the scatter plots.

```{r}
#| label: linear-2c-1
#| warning: false
#| echo: true

# pef2 vs pef 1
lm_pef2_pef1 <- lm(pefsit2 ~ pefsit1, data = lung_data)
summary(lm_pef2_pef1)
```



```{r}
#| label: linear-2c-2
#| warning: false
#| echo: true

# pef1 vs weight 
lm_pef1_weight <- lm(pefsit1 ~ weight, data = lung_data)
summary(lm_pef1_weight)
```

#### 2d) 

Make residual analysis for the analyses you did before. Interpret the results.

```{r}
#| label: linear-2d-1
#| warning: false
#| echo: true

par(mfrow = c(2, 2)) # plot 2 by 2
plot(lm_pef2_pef1)

plot(lm_pef1_weight)
```

#### 2e)

Make a regression analysis with `pefsit1` as dependent variable, and `sex` and `weight` as independent variables. Assess the model fit. Interpret the results.

```{r}
#| label: linear-2e-1
#| warning: false
#| echo: true

# pefsit1 vs (weight, gender)
# note that we converted gender into categorical
lm_pef1_weight_gender <- lm(pefsit1 ~ weight + gender, 
                            data = lung_data)
# lm_pef1_weight_gender
summary(lm_pef1_weight_gender)
```

#### 2f)

Make a regression analysis with `pefmean` as dependent variable, and try out **combinations** of `sex`, `height`, `weight` as independent variables.

Which variables would you include in your final analysis? How much variation is explained? Assess the model fit, and interpret the results. 

```{r}
#| label: linear-2f-1
#| warning: false
#| echo: true

# we can try two sets
# 1. height weight gender
lm_pefm_height_weight_gen <- lm(pefmean ~ height + weight + gender, 
                                data = lung_data)

summary(lm_pefm_height_weight_gen)
```


```{r}
#| label: linear-2f-2
#| warning: false
#| echo: true

# 2. weight gender
lm_pefm_weight_gen <- lm(pefmean ~ weight + gender, 
                                data = lung_data)

summary(lm_pefm_weight_gen)
```


## Exercise 3 (length of hospital stay)

The data was collected at the Geriatric Department at Ullevål Sykehus. Below is a description of the data set `liggetid`. The file includes the following variables:

- Year of birth (faar)
- Month of birth (fmaan)
- Day of birth (fdag)
- Year of hospital admission (innaar)
- Month of admission (innmaan)
- Day of admission (inndag)
- Year of discharge from hospital (utaar)
- Month of discharge (utmaan)
- Day of discharge (utdag)
- Gender, where 1 = male and 0 = female (kjoenn)
- Admission from, where 1 = home, 2 = Div. of Medicine, 3 =
Div. of Surgery, 4 = Other division, 5 = Other hospital, 6 =
Nursing home (kom_fra)
- Stroke, where 1 = yes, 0 = no (slag)
- Age (alder)
- Hospital stay, in days (liggetid)
- Logarithm of hospital stay (lnliggti)
- Comes from Div. of Medicine (kom_fra2)
- Comes from Div. of Surgery (kom_fra3)
- Comes from Other division (kom_fra4)
- Comes from Other hospital (kom_fra5)
- Comes from Nursing home (kom_fra6)
- Censoring variable (censor)

The variable `liggetid` time is calculated from the `innaar`, `innmaan`, `inndag`, `utaar`, `utmaan` and `utdag` variables. 

#### 3a)

Create a box plot for length of hospital stay for men and women.

```{r}
#| label: linear-3a-1
#| warning: false
#| echo: true

liggetid <- read.csv('data/liggetid.csv')
head(liggetid, 3)

# boxplot
boxplot(liggetid ~ kjoenn, data = liggetid)
```


#### 3b) 

We want to explain the variation in lengths of hospital stay. We will look at the independent variables `kjoenn` (gender) and `slag` (stroke). 

Run a regression analysis using the dependent variable `liggetid`. Also perform a residual analysis. What do you think about this analysis?

```{r}
#| label: linear-3b-1
#| warning: false
#| echo: true

# response (dep): liggetid
# predictor (indep): kjoenn, slag

lm_ligge <- lm(liggetid ~ slag + kjoenn, 
               data = liggetid)
summary(lm_ligge)
```

```{r}
#| label: linear-3b-2
#| warning: false
#| echo: true

# visualize the residual
par(mfrow = c(2, 2))
plot(lm_ligge)
```


#### 3c) 

Do the same analysis, but on the log-transformed data. The transformed variable already exists in the dataset, `lnliggti`. Comment on the results.

```{r}
#| label: linear-3c-1
#| warning: false
#| echo: true

# response (dep): log transformed (lnliggti)
# predictor (indep): kjoenn, slag

lm_logligge <- lm(lnliggti ~ slag + kjoenn, 
                  data = liggetid)
summary(lm_logligge)
```


```{r}
#| label: linear-3c-2
#| warning: false
#| echo: true

par(mfrow = c(2, 2))
plot(lm_logligge)
```