---
title: "Survival analysis"
description: "Kaplan-Meier curve, log rank tests"
format: 
  html:
    code-fold: false
    code-tools: true
editor: source
---

# Melanoma 

(This is the example 2 in the exercises)

First we load the dataset, and check what columns are there.

```{r}
#| label: melanoma-1
#| warning: false
#| echo: true
library(survival)

melanoma <- haven::read_dta('data/melanoma.dta')

head(melanoma)
colnames(melanoma)
```

### a) Recode variable

We only need "death from melanoma" or else, hence we need to recode the `status` variable which originally had three categories.

```{r}
#| label: melanoma-a1
#| warning: false
#| echo: true

# if status == 1, code 1; otherwise, code 0
# this indicates whether deaths from melanoma
death <- ifelse(melanoma$status == 1, 1, 0)

# check if the labels are correct
table(death)
table(melanoma$status)

```

You can see that the category `melanoma$status == 1` should match `death == 1`, and `melanoma$status` of 2 and 4 are merged together and coded as `death == 0`. 

### b) Kaplan-Meier plot 


```{r}
#| label: melanoma-b1
#| warning: false
#| echo: true

# take lifetime variable
lifetime <- melanoma$lifetime


km_fit <- survfit(Surv(lifetime, death) ~ 1)
plot(km_fit)

# add title and text
title(main = 'Kaplan-Meier survival estimate', 
      xlab = 'Time', 
      ylab = 'Survival probability')

```


Now we check only time 1, 2, 5. 

```{r}
#| label: melanoma-b2
#| warning: false
#| echo: true

# time 1, 2, 5
tme <- c(1, 2, 5)
summary(km_fit, times = tme)
```


### c) Compare two genders 

```{r}
#| label: melanoma-c1
#| warning: false
#| echo: true

# gender == 1: female; gender == 2: male
gender <- melanoma$gender


km_fit_gender <- survfit(Surv(lifetime, death) ~ gender)
plot(km_fit_gender, col = c('blue', 'red'))
title(main = 'Kaplan-Meier survival estimate: stratify by gender', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'bottomleft', 
       legend = c('Female', 'Male'), 
       lty = c('solid', 'solid'), 
       col = c('blue','red'))

```

Now we test the difference with log-rank test

```{r}
#| label: melanoma-c2
#| warning: false
#| echo: true

# test difference with logrank test 
survdiff(Surv(lifetime, death) ~ gender)
```


### d) Compare tumor thickness

First we do some data processing

```{r}
#| label: melanoma-d1
#| warning: false
#| echo: true

# check the original data type 
melanoma$grouped_tumor_thickness

# you can code them so that the data is more readable
grouped_tumor_thickness <- factor(melanoma$grouped_tumor_thickness, 
                           levels = c('1','2', '3'),
                           labels = c('0-2 mm', '2-5 mm', '5+ mm'))

# check if it is coded correctly
table(grouped_tumor_thickness)
table(melanoma$grouped_tumor_thickness)

```


Now we plot the KM plot, and carry out a log-rank test
```{r}
#| label: melanoma-d2
#| warning: false
#| echo: true

km_fit_tumor <- survfit(Surv(lifetime, death) ~ grouped_tumor_thickness)
plot(km_fit_tumor, col = c('blue', 'red', 'forestgreen'))

title(main = 'Kaplan-Meier survival estimate: stratify by tumor thickness', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'bottomleft', 
        legend = c('0-2 mm', '2-5 mm', '5+ mm'), 
        lty = c('solid', 'solid', 'solid'), 
        col = c('blue','red', 'forestgreen'))


# test difference with logrank test 
survdiff(Surv(lifetime, death) ~ grouped_tumor_thickness)

```




# Length of hospital stay 

(This is the exercise 3 from the exercise for day 7)


```{r}
#| label: liggetid
#| warning: false
#| echo: true

liggetid <- haven::read_dta('data/liggetid.dta')
head(liggetid)
colnames(liggetid)

# all status are 1 
status <- liggetid$status
```


### a) Length of stay vs gender

```{r}
#| label: liggetid-a1
#| warning: false
#| echo: true

# length of stay (los) vs gender
los <- liggetid$liggetid
head(los)

# can choose to recode; or leave as it is;
# gender == 0: female; gender == 1: male
gender <- liggetid$kjoenn
head(gender)

# fit km 
km_liggetid_gender <- survfit(Surv(los, status) ~ gender)

plot(km_liggetid_gender, col = c('blue', 'red'))

title(main = 'Kaplan-Meier survival estimate: stratify by gender', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'topright', 
        legend = c('Female', 'Male'), 
        lty = c('solid', 'solid'), 
        col = c('blue','red'))

```

Now we do log-rank test

```{r}
#| label: liggetid-a2
#| warning: false
#| echo: true

# log rank test 
survdiff(Surv(los, status) ~ gender)
```


### b) Length of stay vs stroke 

```{r}
#| label: liggetid-b1
#| warning: false
#| echo: true


# length of stay vs stroke
# slag == 1: yes, slag == 2: no
stroke <- liggetid$slag

# fit km 
km_liggetid_stroke <- survfit(Surv(los, status) ~ stroke)

plot(km_liggetid_stroke, col = c('blue', 'red'))

title(main = 'Kaplan-Meier survival estimate: stratify by stroke', 
      xlab = 'Time', 
      ylab = 'Survial probability')

# add legend
legend( 'topright', 
        legend = c('Stroke: yes', 'Stroke: no'), 
        lty = c('solid', 'solid'), 
        col = c('blue','red'))

```


```{r}
#| label: liggetid-b2
#| warning: false
#| echo: true

# log rank test
survdiff(Surv(los, status) ~ stroke)
```