---
title: "Logistic regression"
description: "Logistic regression"
format: 
  html:
    code-fold: false
    code-tools: true
editor: source
---

Datasets

* Exercise 1: `birth` ([rda link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/birth.rda), [csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/birth.csv))
* Exercise 2: `framingham` ([rda link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/framingham.rda), [csv link](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/data/framingham.csv))



[R Script](https://github.com/ocbe-uio/teaching_mf9130e/blob/main/lab/code/9_logisticreg.R)

------------------


# Examples (exercises with solution)

## Exercise 1 (birth)

We shall analyse the data set given in the file `birth.dta`. In a study in Massachusetts, USA, birth weight was measured for the children of 189 women. The main variable in the study was birth weight, `BWT`, which is an important indicator of the condition of a newborn child. Low birth weight (below 2500 g) may be a medical risk factor. A major question is whether smoking during pregnancy influences the birth weight. One has also studied whether a number of other factors are related to birth weight, such as hypertension in the mother. 

The variables of the study are

* Identification number (ID)
* Low birth weight (LOW)
* Age of the mother (AGE)
* Weight (in pounds) at last menstruaal period (LWT)
* Ethnicity, 1 means White, 2 means African American, 3 meaans other (ETH)
* Smoking status, 1 means current smoker, 0 means not smoking during pregnancy (SMK)
* History of premature labbour, values 0, 1, 2... (PTL)
* History of hypertension, 1 is yes, 0 is no (HT)
* Uterine irritability, 1 is yes, 0 is no (UI)
* First trimester visits, values 0, 1, 2... (FTV)
* Third trimster visits, values 0, 1, 2... (TTV)
* Birth weight (BWT)


```{r}
#| label: lr-1a-1
#| warning: false
#| echo: true

# load data
birth_data <- read.csv('data/birth.csv')

# print the first rows of the data set
head(birth_data)
```

### 1a)

Perform a logistic regression analysis with `low` as dependent variable, and `age` as independent variable. Interpret the results.

::: {.callout-note}
## Recode `low` as `is_low_bwt`

With logistic regression (glm()), you need to specify the outcome variable as 

* either 0 or 1 (numeric values)
* or a factor (ordered so that it matches 0 or 1 category)

In the example we should the first option (numeric 0 or 1)

It could be a good practice to create another variable, `is_low_bwt`, so that you keep the original `low` variable untouched; this is useful in case you want to double check if the coding is correct or need to start over.

:::



```{r}
#| label: lr-1a-2
#| warning: false
#| echo: true

# be careful with the levels: bwt<= 2500 is the cases (code as 1)
table(birth_data$low)

# recode low variable (numeric 1 and 0)
birth_data$is_low_bwt <- ifelse(birth_data$low == 'bwt <= 2500',1,0)

# check each category
table(birth_data$is_low_bwt)
```


```{r}
#| label: lr-1a-3
#| warning: false
#| echo: true

# fit lr (using is_low_bwt)
lr_low_age <- glm(is_low_bwt ~ age, 
                  data = birth_data, 
                  family = 'binomial')
summary(lr_low_age)

# double check whether 0/1 categories are correct
table(lr_low_age$y)
```

```{r}
#| label: lr-1a-4
#| warning: false
#| echo: true

# the result regresssion coefficient is not odds ratio
lr_low_age$coefficients

# equivalently,
# coefficients(lr_low_age)
# coef(lr_low_age)
# confidence interval for beta:
# confint(lr_low_age)


# to get odds ratio, exponentiate 
exp(coef(lr_low_age)) # OR
exp(confint(lr_low_age)) # CI for OR
```




### 1b) 

Perform a logistic regression analysis with `low` as dependent variable and `smk` as independent variable. 

```{r}
#| label: lr-1b-1
#| warning: false
#| echo: true

# low ~ smk 
lr_low_smk <- glm(is_low_bwt ~ smk, 
                  data = birth_data, 
                  family = 'binomial')
summary(lr_low_smk)

# odds ratio 
exp(coef(lr_low_smk))
exp(confint(lr_low_smk))
```

### 1c) 

Perform a logistic regression with `low` as dependent variable, and `eth` as independent variable. Be careful with the reference category.


```{r}
#| label: lr-1c-1
#| warning: false
#| echo: true


# low ~ eth 
# ethnicity needs to be factorised
# baseline: white
birth_data$eth <- eth <- factor(birth_data$eth, 
              levels = c('white', 'black', 'other'), 
              labels = c('white', 'black', 'other'))
head(birth_data$eth)

# fit lr 
lr_low_eth <- glm(is_low_bwt ~ eth, 
                  data = birth_data, 
                  family = 'binomial')
summary(lr_low_eth)

# odds ratio 
exp(coef(lr_low_eth))
exp(confint(lr_low_eth))

```



### 1d)

Perform a logistic regression with `low` as dependent variable, and `age`,`smk` and `eth` as independent variables.



```{r}
#| label: lr-1d-1
#| warning: false
#| echo: true

lr_low_all <- glm(is_low_bwt ~ age + smk + eth, 
                  data = birth_data, 
                  family = 'binomial')
summary(lr_low_all)

# odds ratio 
exp(coef(lr_low_all))
exp(confint(lr_low_all))
```

### 1e) 

Based on the above analysis, set up a result table which reports:

* odds ratios OR (unadjusted, and adjusted)
* 95% confidence intervals for OR
* p-values for OR

Make sure you know how to interpret the table. 


## Exercise 2 (framingham)

We use the data from the Framingham study, `framingham`. The dataset contains a selection of n = 500 men aged 31 to 65 years. 

The response variable is FIRSTCHD, and this is equal to 1 if the individual has coronary heart disease and 0 otherwise. 

We have four explanatory variables: 

* MEANSBP, the average systolic blood pressure (mmHg) of two blood pressure measurements; 
* SMOKE which is smoking (1 = yes, 0 = no);
* CHOLESTEROL which is serum cholesterol in mg/dl;
* AGE (age in years). 


### 2a)

Analyse the relationship between `firstchd` and `smoke` in a logistic regression model.

```{r}
#| label: lr-2a-1
#| warning: false
#| echo: true

framingham <- read.csv('data/framingham.csv')
head(framingham)

# check the outcome variable
# evidence is 1, no-evidence is 0
table(framingham$firstchd)

# code into 1 and 0 (numeric)
framingham$firstchd_coded <- ifelse(framingham$firstchd == 'evidence', 1, 0)

# check if correct
table(framingham$firstchd_coded)
```

```{r}
#| label: lr-2a-2
#| warning: false
#| echo: true

# firstchd, smoke 
lr_chd_smk <- glm(firstchd_coded ~ smoke, 
                  data = framingham, 
                  family = 'binomial')
# double check if category is correct
table(lr_chd_smk$y)  

summary(lr_chd_smk)

# odds ratio 
exp(coef(lr_chd_smk))
exp(confint(lr_chd_smk))
```


### 2b)

Analyse the relationship between `firstchd` and `meansbp` in a logistic regression model.

```{r}
#| label: lr-2b-1
#| warning: false
#| echo: true

# firstchd, meansbp
lr_chd_bp <- glm(firstchd_coded ~ meansbp, 
                 data = framingham, 
                 family = 'binomial')

summary(lr_chd_bp)

# odds ratio 
exp(coef(lr_chd_bp))
exp(confint(lr_chd_bp))
```

### 2c)

Include also the other two explanatory variables in a logisic regression model. Interpret the results. 

```{r}
#| label: lr-2c-1
#| warning: false
#| echo: true

lr_chd_all <- glm(firstchd_coded ~ meansbp + smoke + cholesterol + age, 
                  data = framingham, 
                  family = 'binomial')
summary(lr_chd_all)

# odds ratio 
round(exp(coef(lr_chd_all)), digits = 3)
round(exp(confint(lr_chd_all)), digits = 3)
```


